Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus

### 2024-07-09 manuscript_thev_transcriptome.Rmd
addressed reviewer comments from virology journal

## ABSTRACT   
**Background:** Hemorrhagic enteritis, caused by _Turkey Hemorrhagic Enteritis Virus (THEV)_, is a disease affecting turkey poults characterized by immunosuppression and bloody diarrhea. An avirulent THEV strain that retains the immunosuppressive ability is used as a live vaccine. Characterizing the splice map of THEV is an essential step that would allow studies of individual genes mediating its immunosuppressive functions. We used RNA sequencing to characterize the splice map of THEV for the first time, providing key insights into the THEV gene expression and mRNA structures.   
**Methods:** After infecting a turkey B-cell line with the vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was extracted and poly-A-tailed mRNA sequenced. Reads were mapped to the THEV genome after trimming and transcripts assembled with StringTie. We performed PCR of THEV cDNA, cloned the PCR products, and used Sanger sequencing to validate all identified splice junctions.   
**Results:** Researchers previously annotated the THEV genome as encoding 23 open reading frames (ORFs). We identified 29 spliced transcripts from our RNA sequencing data, all containing novel exons although some exons matched some previously annotated ORFs. The three annotated splice junctions were also corroborated by our data. During validation we identified five additional unique transcripts, a subset of which were further validated by 3' rapid amplification of cDNA ends (3' RACE). Thus, we report that the genome of THEV contains 34 transcripts with the coding capacity for all annotated ORFs. However, we found six of the previously annotated ORFs to be truncated ORFs on the basis of the identification of an in-frame upstream start codon or the detection of additional coding exons. We also identified three of the annotated ORFs with longer or shorter isoforms, and seven novel unannotated ORFs that could potentially be translated; although it is beyond the scope of this manuscript to investigate whether they are translated.   
**Conclusions:** Similar to human adenoviruses, all THEV transcripts are spliced and organized into five transcription units under the control of their cognate promoters. The genes are expressed under temporal regulation and THEV also produces multiple distinctly spliced transcripts that code for the same protein.

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all types of vertebrates studied to date. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via highly complex alternative splicing patterns [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs, presumably inherited from a common ancestor. All other genes are termed "genus-specific". The genus-specific genes tend to be located at the termini of the genome while genus-common genes are usually towards the center of the genome [@Davison2003]. This pattern is also observed in *Poxviridae* and *Herpesviridae*, which also have linear DNA genomes [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) belongs [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009; @Guimet2016]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content of all known AdVs (see **Figure 1**) [@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a debilitating acute disease predominantly affecting turkey poults characterized by immunosuppression, intestinal lesions leading to bloody diarrhea, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967; @Harrach2008]. While the current vaccine strain (a THEV-A called Virginia Avirulent Strain [VAS]) has proven effective at preventing HE in turkey poults, it still retains its immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated birds leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. To eliminate the immunosupressive effect of the vaccine strain, a thorough investigation of the culprit viral genes mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making an investigation of specific immunosuppressive viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current AdV gene expression and replication knowledge is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV transcription is temporally regulated; their genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late transcription unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host-cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, or direct genome packaging. The immediate early genes E1A are expressed first, followed by the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. It is noteworthy that the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but it reaches its maximal activity during late infection (after DNA replication). However, during early infection only a subset of the MLP-derived transcripts are expressed [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all containing three non-coding exons at the 5'-end (collectively known as the tripartite leader; TPL) [@Donovan2020; @Zhao2014]. There is also an alternate three-exon 5' non-coding leader sequence present in varying amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 gene products, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies using long-read direct RNA sequencing strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Overview of sequencing data and analysis pipeline outputs**   
A prior study by Aboezz *et al*. demonstrated that nearly all THEV transcripts became detectable starting at 4 hours post-infection (hpi), with one replication cycle concluding around 18 hpi [@Aboezz2019]. Consequently, we harvested infected MDTC-RP19 cells (MOI of 100 genome copy numbers/cell) at 4-, 12-, 24-, and 72-hpi to capture all transcripts within a broad time window. Our paired-end RNA sequencing (RNA-seq) experiment generated an average of `r avg_tot_count` million total reads of `r seq_len` bp length per time-point. These reads were concurrently mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the Hisat2 [@Pertea2016] reference-based aligner. A total of `r total_mapped` million reads from all time-points mapped to the virus genome, providing comprehensive coverage and leaving no regions unmapped. The mapped reads to the virus genome increased significantly from a scant `r cov_4hr` reads at 4 hpi to `r cov_72hr` million reads at 72 hpi (**Table 1**, **Figure 2A**). From these mapped reads, we identified `r comma(nrow(total_unq_jncs_submit))` unique THEV splice junctions across all time-points, with later time-points exhibiting significantly more sequence reads supporting the splice junctions than earlier time-points. For instance, all `r tot_jun_4` unique junctions at 4 hpi had fewer than 10 supporting reads each, averaging only `r mean_juncreads_4` reads per junction. In contrast, the `r tot_jun_72` unique junctions at 72 hpi averaged `r mean_juncreads_72` reads per junction, with some junctions reaching as high as `r max_junread72` reads. The marked increase in splice junction and mapping reads to the THEV genome over time indicates an active infection and successful viral replication, which is corroborated by our quantitative PCR (qPCR) assay that quantified the total number of viral genome copies over time (**Figure 2B**).

@@

The E2B region transcripts also start with the E2-5'UTR but extend downstream to reach the TTS at 2334 bp in the IM region, which is in the immediate context of a canonical PAS (position 2333 bp) where polyadenylation likely occurs. The E2B transcripts, TRXPT_6 and TRXPT_7, are almost identical except for an extra splice junction at the 3'-end of TRXPT_6 (**Figure 7**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, with the pTP ORF predicted to be spliced from two exons. The predicted splice junction of pTP is corroborated by our data but the full transcript is markedly longer than the predicted ORF, although the encoded product (pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of pTP with secondary start codon (secSC) usage. The CP of TRXPT_6 slightly differs from TRXPT_7 because a new STC resulting from the extra splice site forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

**Major Late Transcription Unit (MLTU) or MLP Region transcripts**   
The MLTU transcripts, dominant in the late phase of the AdV infectious cycle, are produced by alternative polyadenylation and splicing of a primary transcript and grouped into five transcript classes (L1-L5). About 13 out of the 23 predicted ORFs in THEV fall within this TU, some of which we have categorized under the E3 TU instead. Our RNA-seq data revealed 12 transcripts (`r paste0("TRXPT_", c(8:14, 16:20), collapse = ", ")`) in this TU, most of which have the 5' TPL sequence as in all AdVs. However, three transcripts (`r paste0("TRXPT_", c(16:18),collapse = ", ")`) use a different leader sequence (sTPL), where a different first exon is used instead of the first TPL exon, and TRXPT_20 uses only the third TPL exon (TPL3); see **Figure 10**.

@@

While the advent of next-generation sequencing has rendered easier the study of large and complex eukaryotic transcriptomes, studying the smaller, compact viral transcriptomes is counterintuitively more challenging, as the transcripts typically have significant overlaps due to genome economization. AdV transcriptomes escalate the difficulty due to the wide array of mRNAs produced via very complex alternative splicing and polyadenylation, all initiated from relatively few promoters. Standard RNA-seq analysis programs, not primarily designed for such compact, gene-dense, and complex transcriptomes, further compound this challenge. Furthermore, in our case, no prior transcriptomic studies for THEV exist. Our approach combines standard RNA-seq analysis programs with custom analyses and experimentally validating all splice junctions with independent methods. The transcript map for THEV produced from our analysis is strikingly similar to that of the MAdVs.

Our work provides the first insights into THEV splicing, revealing 34 transcripts grouped into five transcription units (TUs). The general temporal gene expression regulation observed in MAdVs, with early regions peaking at earlier time points followed by MLTU predominance at later time points, seems to also apply to THEV. An unexpected observation is that the pileup of mapped reads to THEV seems consistently skewed over similar regions of the genome at all time points. Given the temporal regulation of AdVs gene expression, we anticipated distinct differences in read pileups over the genome at different time points, indicating the different stages of infection. This may be due to an unsynchronized infection, leading to transcripts overlapping the time points. Further research is required to determine the precise temporal regulation of THEV.

Short read deep sequencing effectively reconstructs full AdV mRNA structures, particularly mapping splice sites [@Zhao2014]. However, the substantial overlapping nature of AdV mRNAs and fragmentation during library preparation make it challenging to map the exact TSS, TTS, and PASs of assembled transcripts. Also, transcripts with significant overlaps may be assembled as a single, longer mRNA, since the short reads alone do not provide enough context for the transcript assembler (StringTie) to distinguish them. Our results show transcripts in the same TU initiating or terminating in similar areas, but not at the exact same position. We consider the most upstream TSS or most downstream TTS for for the transcripts involved but we present them unchanged in all the figures shown (see **Supplementary Table S2**). Also, comparing our results to the better-studied MAdV transcriptomes, we believe some long transcripts in the MLTU (TRXPT_9, TRXPT_10, and TRXPT_11) are likely due to fusing some E3 transcripts to the terminal exons of the MLTU transcripts by StringTie, making them significantly longer. These mRNAs have unusually many exons and their last few exons are identical to some E3 mRNAs. Future studies using long read sequencing technologies will provide more precise mapping of the TSS and TTS and clarify the structures of the long MLTU transcripts.

While most predicted ORFs are encoded by the spliced transcripts, we found some that seem to be truncated predictions, as either an upstream in-frame SC or unknown upstream exons were found. Other ORFs were identified that were either shorter or longer isoforms of some predicted ORF. We also found several novel unpredicted ORFs (**Supplementary Table S3**). On this basis, we anticipate that further studies will likely reveal more unpredicted novel ORFs or new variants of predicted ORFs. Furthermore, it is not unreasonable to presume that several splice variants will likely be found as evidenced firstly by finding unique transcripts using 3' RACE and during our splice junction validation steps. And secondly, recent studies [@Donovan2020; @Zhao2014; @Westergren2021] are still discovering novel mRNA variants for even the best studied MAdVs decades later.

Eukaryotic mRNAs are typically functionally monocistronic, with the 5’-most AUG determining the translation reading frame. However, AdV mRNAs, which span more than one ORF, are functionally polycistronic, employing non-standard mechanisms of translation initiation such as secSC usage and ribosome shunting [@Westergren2021; @Guimet2016]. AdVs use secondary AUGs as initiation codons for most E1b proteins and some E3 proteins.  In fact, recent studies show that secSC usage is found transcriptome-wide. This is thought to occur because translation initiation at the first SC is inefficient, allowing downstream SCs to be employed [@Donovan2020]. Ribosomal shunting or jumping mechanism is utilized for MLTU transcripts that have the TPL. This mechanism allows the ribosome to translocate to a downstream AUG, under the direction of the shunting elements in the TPL, even if a start codon in a good Kozak sequence context is bypassed. Thus, predicting the protein(s) expressed from an AdV mRNA is uncertain as any one of the AUGs may be selected [@Guimet2016; @Westergren2021]. Almost all the THEV transcripts in our data have the CP for several ORFs, some spanning as many as six ORFs. This supports the usage of these special ribosome initiation mechanisms as several predicted and novel ORFs found on mRNA in our data could not be translated using only the typical ribosome scanning mechanism. Interestingly, several distinct THEV mRNAs have identical CPs. This is also observed in human AdVs in a recent study [@Donovan2020]. They proposed that this may permit protein production to be fine-tuned through alteration in the balance between different mRNA groups expressing that ORF.

@@

The THEV transcriptome bears remarkable similarity to the better-studied MAdVs. The transcriptome is organized into five TUs, with temporal regulation divided into early and late genes, and a broad repertoire of transcripts are produced via virtually unlimited alternative splicing. However, the THEV transcriptome appears less sophisticated (i.e, it encodes fewer genes) than MAdVs, primarily because the MAdV genomes are close to twice as long as that of THEV. The lack of subdivision of the E1 region into E1a and E1b is one of the most obvious examples. Also, the MAdV E4 region encodes several proteins unlike in THEV where only one transcript encoding one protein was found. The complexity of the MLTU leader sequences is another example. While the majority of the THEV MLTU transcripts begin with the TPL just like MAdVs with a small subset using a variant leader sequence (sTPL), significantly more diverse 5'-UTRs are employed for MAdV MLTU transcripts. Namely, the TPL, the so-called x, y, and z leaders, and the i-leader are 5' leaders utilized by MAdV MLTU mRNAs. The absence of these non-TPL leaders in our data could mean that the 5'-UTR diversity of THEV's MLTU mRNAs is more limited due to its smaller genome size or future studies could uncover more variety not seen in our results. Being the first transcriptomic characterization of THEV, this work should serve as useful resource to future THEV gene expression and transcriptomic studies, especially, mapping the mRNA splice sites.
\newpage

>>>>

## ABSTRACT   
**Background:** Hemorrhagic enteritis, caused by _Turkey Hemorrhagic Enteritis Virus (THEV)_, is a disease affecting turkey poults characterized by immunosuppression and bloody diarrhea. An avirulent THEV strain that retains the immunosuppressive ability is used as a live vaccine. Characterizing the splice map of THEV is an essential step that would allow studies of individual genes mediating its immunosuppressive functions. We used RNA sequencing to characterize the splice map of THEV for the first time, providing key insights into the THEV gene expression and mRNA structures.   
**Methods:** After infecting a turkey B-cell line with the vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was extracted, and poly-A-tailed mRNA sequenced. Reads were mapped to the THEV genome after trimming and transcripts assembled with StringTie. We performed PCR of THEV cDNA, cloned the PCR products, and used Sanger sequencing to validate all identified splice junctions.   
**Results:** Researchers previously annotated the THEV genome as encoding 23 open reading frames (ORFs). We identified 29 spliced transcripts from our RNA sequencing data, all containing novel exons although some exons matched some previously annotated ORFs. The three annotated splice junctions were also corroborated by our data. During validation we identified five additional unique transcripts, a subset of which were further validated by 3' rapid amplification of cDNA ends (3' RACE). Thus, we report that the genome of THEV contains 34 transcripts with the coding capacity for all annotated ORFs. However, we found six of the previously annotated ORFs to be truncated ORFs on the basis of the identification of an in-frame upstream start codon or the detection of additional coding exons. We also identified three of the annotated ORFs with longer or shorter isoforms, and seven novel unannotated ORFs that could potentially be translated; although it is beyond the scope of this manuscript to investigate whether they are translated.   
**Conclusions:** Similar to human adenoviruses, all THEV transcripts are spliced and organized into five transcription units under the control of their cognate promoters. The genes are expressed under temporal regulation and THEV also produces multiple distinctly spliced transcripts that code for the same protein. Studies of the newly identified potential proteins should be urgently performed as these proteins may have roles in THEV-induced immunosuppression. Also, knowing the splicing of THEV genes should be invaluable to future research focusing of studying THEV genes, as this will allow accurate cloning of the mRNAs.

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all types of vertebrates studied to date. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via highly complex alternative splicing patterns [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs, presumably inherited from a common ancestor. All other genes are termed "genus-specific". The genus-specific genes tend to be located at the termini of the genome while genus-common genes are usually towards the center of the genome [@Davison2003]. This pattern is also observed in *Poxviridae* and *Herpesviridae*, which also have linear DNA genomes [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) belongs [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009; @Guimet2016]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content of all known AdVs, with five genus-specific genes of undefined functions (see **Figure 1**) [@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a debilitating acute disease predominantly affecting turkey poults characterized by immunosuppression, intestinal lesions leading to bloody diarrhea, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967; @Harrach2008]. While the current vaccine strain (a THEV-A called Virginia Avirulent Strain [VAS]) has proven effective at preventing HE in turkey poults, it still retains its immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated birds leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. To eliminate the immunosuppressive effect of the vaccine strain, a thorough investigation of the culprit viral genes mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making an investigation of specific immunosuppressive viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current AdV gene expression and replication knowledge is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV transcription is temporally regulated; their genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late transcription unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host-cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, or direct genome packaging. The immediate early genes E1A are expressed first, followed by the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. It is noteworthy that the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but it reaches its maximal activity during late infection (after DNA replication). However, during early infection only a subset of the MLP-derived transcripts are expressed [@Guimet2016]. MAdVs make extensive use of alternative RNA splicing and polyadenylation to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all containing three non-coding exons at the 5'-end (collectively known as the tripartite leader; TPL) [@Donovan2020; @Zhao2014]. There is also an alternate three-exon 5' non-coding leader sequence present in varying amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 gene products, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies using long-read direct RNA sequencing strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Overview of sequencing data and analysis pipeline outputs**   
A prior study by Aboezz *et al*. demonstrated that nearly all THEV transcripts became detectable starting at 4 hours post-infection (hpi), with one replication cycle concluding around 18 hpi [@Aboezz2019]. Consequently, we harvested infected MDTC-RP19 cells (multiplicity of infection (MOI) of 100 genome copy numbers/cell) at 4-, 12-, 24-, and 72-hpi to capture all transcripts within a broad time window. Our paired-end RNA sequencing (RNA-seq) experiment generated an average of `r avg_tot_count` million total reads of `r seq_len` bp length per time-point. These reads were concurrently mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the Hisat2 [@Pertea2016] reference-based aligner. A total of `r total_mapped` million reads from all time-points mapped to the virus genome, providing comprehensive coverage and leaving no regions unmapped. The mapped reads to the virus genome increased significantly from a scant `r cov_4hr` reads at 4 hpi to `r cov_72hr` million reads at 72 hpi (**Table 1**, **Figure 2A**). From these mapped reads, we identified `r comma(nrow(total_unq_jncs_submit))` unique THEV splice junctions across all time-points, with later time-points exhibiting significantly more sequence reads supporting the splice junctions than earlier time-points. For instance, all `r tot_jun_4` unique junctions at 4 hpi had fewer than 10 supporting reads each, averaging only `r mean_juncreads_4` reads per junction. In contrast, the `r tot_jun_72` unique junctions at 72 hpi averaged `r mean_juncreads_72` reads per junction, with some junctions reaching as high as `r max_junread72` reads. The marked increase in splice junction and mapping reads to the THEV genome over time indicates an active infection and successful viral replication, which is corroborated by our quantitative PCR (qPCR) assay that quantified the total number of viral genome copies over time (**Figure 2B**).

@@

The E2B region transcripts also start with the E2-5'UTR but extend downstream to reach the TTS at 2334 bp in the IM region, which is in the immediate context of a canonical PAS (position 2333 bp) where polyadenylation likely occurs. The E2B transcripts, TRXPT_6 and TRXPT_7, are almost identical except for an extra splice junction at the 3'-end of TRXPT_6 (**Figure 7**). TRXPT_7 has the CP for both classical proteins (pTP and Ad-pol) encoded in this region, with the pTP ORF predicted to be spliced from two exons. The predicted splice junction of pTP is corroborated by our data but the full transcript is markedly longer than the predicted ORF, although the encoded product (pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of pTP with secondary SC (secSC) usage. The CP of TRXPT_6 slightly differs from TRXPT_7 because a new STC resulting from the extra splice site forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

**MLTU or MLP Region transcripts**   
The MLTU transcripts, dominant in the late phase of the AdV infectious cycle, are produced by alternative polyadenylation and splicing of a primary transcript and grouped into five transcript classes (L1-L5). About 13 out of the 23 predicted ORFs in THEV fall within this TU, some of which we have categorized under the E3 TU instead. Our RNA-seq data revealed 12 transcripts (`r paste0("TRXPT_", c(8:14, 16:20), collapse = ", ")`) in this TU, most of which have the 5' TPL sequence as in all AdVs. However, three transcripts (`r paste0("TRXPT_", c(16:18),collapse = ", ")`) use a different leader sequence (sTPL), where a different first exon is used instead of the first TPL exon, and TRXPT_20 uses only the third TPL exon (TPL3); see **Figure 10**.

@@

While the advent of next-generation sequencing has rendered easier the study of large and complex eukaryotic transcriptomes, studying the smaller, compact viral transcriptomes is counterintuitively more challenging, as the transcripts typically have significant overlaps due to genome economization. AdV transcriptomes escalate the difficulty due to the wide array of mRNAs produced via very complex alternative splicing and polyadenylation, all initiated from relatively few promoters. Standard RNA-seq analysis programs, not primarily designed for such compact, gene-dense, and complex transcriptomes, further compound this challenge. Furthermore, in our case, no prior transcriptomic studies for THEV exist; hence, assembling the transcripts without any prior experimentally-derived annotation of THEV splicing using only short illumina reads proved difficult. Lastly, we had initially planned to sequence RNA from another time point (8 hpi); however, all the RNA samples from 8 hpi and one replicate sample from 12 hpi got too degraded during the library preparation steps to be yield any useful data. We believe that these would have contributed to better insights into the temporal expression levels and splicing. Our approach combines standard RNA-seq analysis programs with custom analyses and experimentally validating all splice junctions with independent methods. The transcript map for THEV produced from our analysis is strikingly similar to that of the MAdVs.

Our work provides the first insights into THEV splicing, revealing 34 transcripts grouped into five TUs. The general temporal gene expression regulation observed in MAdVs, with early regions peaking at earlier time points followed by MLTU predominance at later time points, seems to also apply to THEV. An unexpected observation is that the pileup of mapped reads to THEV seems consistently skewed over similar regions of the genome at all time points. Given the temporal regulation of AdVs gene expression, we anticipated distinct differences in read pileups over the genome at different time points, indicating the different stages of infection. This may be due to an unsynchronized infection, leading to transcripts overlapping the time points. Further research is required to determine the precise temporal regulation of THEV.

Short read deep sequencing effectively reconstructs full AdV mRNA structures, particularly mapping splice sites [@Zhao2014]. However, the substantial overlapping nature of AdV mRNAs and fragmentation during library preparation make it challenging to map the exact TSS, TTS, and PASs of assembled transcripts. As AdVs make heavy use of alternative polyadenylation, short read RNA-seq is ill-equipped to discriminate mRNA variants of the same gene produced via alternative polyadenylation. Thus, shorter variants of alternatively polyadenylated mRNAs may potentially be incorporated into the longer variants during transcript assembly, significantly diminishing the diversity of mRNA in the transcriptome. Also, independent transcripts with significant overlaps may be assembled as a single, longer mRNA, since the short reads alone do not provide enough context for the transcript assembler (StringTie) to distinguish them. Such fusions may affect the transcript expression level estimations by inflating or deflating the expression levels of the transcripts involved, affecting the proper understanding of the temporal gene expression regulation and also the diversity of the transcriptome. Transcripts that have reads mistakenly fused with them would have inflated expression levels while those whose reads are counted elsewhere would show false lower expression levels. In our case, since we used other independent methods to validate the splice junctions, we believe these drawbacks to be minimized. Our results show transcripts in the same TU initiating or terminating in similar areas, but not at the exact same position. We consider the most upstream TSS or most downstream TTS for for the transcripts involved but we present them unchanged in all the figures shown (see **Supplementary Table S2**). Also, comparing our results to the better-studied MAdV transcriptomes, we believe some long transcripts in the MLTU (TRXPT_9, TRXPT_10, and TRXPT_11) are likely due to fusing some E3 transcripts to the terminal exons of the MLTU transcripts by StringTie, making them significantly longer. These mRNAs have unusually many exons and their last few exons are identical to some E3 mRNAs. Future studies using long read sequencing technologies will provide more precise mapping of the TSS, TTS, and PASs and clarify the structures of the long MLTU transcripts.

While most predicted ORFs are encoded by the spliced transcripts, we found some that seem to be truncated predictions, as either an upstream in-frame SC or unknown upstream exons were found. Other ORFs were identified that were either shorter or longer isoforms of some predicted ORF. We also found several novel unpredicted ORFs (**Supplementary Table S3**). On this basis, we anticipate that further studies will likely reveal more unpredicted novel ORFs or new variants of predicted ORFs. Furthermore, it is not unreasonable to presume that several splice variants will likely be found as evidenced firstly by finding unique transcripts using 3' RACE and during our splice junction validation steps. And secondly, recent studies [@Donovan2020; @Zhao2014; @Westergren2021] are still discovering novel mRNA variants for even the best studied MAdVs decades later. These new potential proteins and isoforms significantly extend the repertoire of THEV gene products than predicted, adding a hefty number of proteins of undefined function to the previously predicted five (See **Figure 1**). These new potential proteins and isoforms of unknown functions may mediate or contribute to viral replication efficiency, or the immunosuppression associated with THEV. Hence, further studies of these potential proteins is urgently needed.

Eukaryotic mRNAs are typically functionally monocistronic, with the 5’-most AUG determining the translation reading frame. However, AdV mRNAs, which span more than one ORF, are functionally polycistronic, employing non-standard mechanisms of translation initiation such as secSC usage and ribosome shunting [@Westergren2021; @Guimet2016]. AdVs use secondary AUGs as initiation codons for most E1b proteins and some E3 proteins.  In fact, recent studies show that secSC usage is found transcriptome-wide. This is thought to occur because translation initiation at the first SC is inefficient, allowing downstream SCs to be employed [@Donovan2020]. Ribosomal shunting or jumping mechanism is utilized for MLTU transcripts that have the TPL. This mechanism allows the ribosome to translocate to a downstream AUG, under the direction of the shunting elements in the TPL, even if a SC in a good Kozak sequence context is bypassed. Thus, predicting the protein(s) expressed from an AdV mRNA is uncertain as any one of the AUGs may be selected [@Guimet2016; @Westergren2021]. Almost all the THEV transcripts in our data have the CP for several ORFs, some spanning as many as six ORFs. This supports the usage of these special ribosome initiation mechanisms as several predicted and novel ORFs found on mRNA in our data could not be translated using only the typical ribosome scanning mechanism. Interestingly, several distinct THEV mRNAs have identical CPs. This is also observed in human AdVs in a recent study [@Donovan2020]. They proposed that this may permit protein production to be fine-tuned through alteration in the balance between different mRNA groups expressing that ORF.

@@

The THEV transcriptome bears remarkable similarity to the better-studied MAdVs. The transcriptome is organized into five TUs, with temporal regulation divided into early and late genes, and a broad repertoire of transcripts are produced via virtually unlimited alternative splicing. However, the THEV transcriptome appears less sophisticated (i.e, it encodes fewer genes) than MAdVs, primarily because the MAdV genomes are close to twice as long as that of THEV. The lack of subdivision of the E1 region into E1a and E1b is one of the most obvious examples. Also, the MAdV E4 region encodes several proteins unlike in THEV where only one transcript encoding one protein was found. The complexity of the MLTU leader sequences is another example. While the majority of the THEV MLTU transcripts begin with the TPL just like MAdVs with a small subset using a variant leader sequence (sTPL), significantly more diverse 5'-UTRs are employed for MAdV MLTU transcripts. Namely, the TPL, the so-called x, y, and z leaders, and the i-leader are 5' leaders utilized by MAdV MLTU mRNAs. The absence of these non-TPL leaders in our data could mean that the 5'-UTR diversity of THEV's MLTU mRNAs is more limited due to its smaller genome size or future studies could uncover more variety not seen in our results. We also note that although THEV genomic sequences show minimal differences between strains [@Beach2009a], the transcriptomes may have significant variations; hence, our results may vary from other THEV strains. Also, performing the study *in vivo* or with primary turkey cells may show different results. The potential new proteins identified in our work adds to the number of proteins with undefined functions in THEV; these may have roles in viral replication efficiency, or the immunosuppression associated with THEV. Hence, further studies of these proteins and other predicted proteins of unknown function should be useful in elucidating THEV-induced immunosuppression. Being the first transcriptomic characterization of THEV, this work should serve as useful resource to future THEV gene expression and transcriptomic studies, especially, mapping the mRNA splice sites. More importantly, this work characterizing the splicing of THEV mRNAs will allow researchers to accurately clone any THEV gene(s) of interest to study its potential role in inducing immunosuppression or other functions.
